\pkg{cv}: An \proglang{R} Package for Cross-Validation of Regression Models

### 2024-01-15 scratch/notes_john/JSS-article/JSS-article-reduced-2.Rmd
compress JSS paper section on contrived data

## Example: CV of clusters versus individuals using contrived data

@@

Figure \ref{fig:plot1} (a) shows a scatterplot the data for a representative group of 10 (without loss of generality, the first 10) of the 100 students, displaying the 95% concentration ellipse for each cluster:

```{r plot1, echo=FALSE, results=FALSE}
#| fig.height = 4,
#| fig.cap = "(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data",
#| fig.subcap=rep("", 5),
#| fig.ncol = 2, out.width = "45%", fig.align = "center"
library("lattice")
library("latticeExtra")
library("lme4", quietly=TRUE)
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot # display graph

@@

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))

@@

The initial mixed-effects model that we fit to the data is a simple random-intercepts model:

```{r}
# random intercept only:
mod.0 <- lmer(y ~ 1 + (1 | group), Data)
summary(mod.0)
```

We will shortly consider three other, more complex, mixed models.

We proceed to obtain predictions from the random-intercept model (\code{mod.0}) for fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students). Predictions for the random-intercept model, \code{mod.0}, for the first 10 students are shown in Figure \ref{fig:plot1} (b). The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean. Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.

Our next model, \code{mod.1}, includes a fixed intercept and fixed effect of \code{x} along with a random intercept:

```{r}
# effect of x and random intercept:
mod.1 <- lmer(y ~ x + (1 | group), Data)
summary(mod.1)
```

Predictions from this model appear in Figure \ref{fig:plot1} (c). The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}. We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod1} to \code{mod0}, but that cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.

Our third model, \code{mod.2}, includes the "contextual effect" of $x$---that is, the cluster mean \code{xm}---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:

```{r}
# effect of x, contextual (student) mean of x, and random intercept:
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
        # equivalent to y ~ I(x - xm) + xm + (1 | group)
summary(mod.2)
```

This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).

@@

Finally, \code{mod.3} directly estimates the model used to generate the data. As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}:

```{r}
# model generating the data (where Bb = 0)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
summary(mod.3)
```

The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar to those from \code{mod.2}.

@@

library("car", quietly=TRUE)
scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")
```
A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors. There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

>>>>

## Example: Contrasting cluster-based and case-based CV

@@

Figure \ref{fig:plot1} (a) shows a scatterplot the data for a representative group of 10 (without loss of generality, the first 10) of the 100 students, displaying the 95% concentration ellipse for each cluster.

```{r plot1, echo=FALSE, results=FALSE}
#| fig.height = 4,
#| fig.cap = "(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data",
#| fig.subcap=rep("", 5),
#| fig.ncol = 2, out.width = "45%", fig.align = "center"
library("lattice")
library("latticeExtra")
library("lme4", quietly=TRUE)
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot # display graph

@@

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=2)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 2), lwd=2)))

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=2)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35) #,
  # key=list(
  #   corner=c(0.95, 0.05),
  #   text=list(c("fixed effects only","fixed and random")),
  #   lines=list(lty=c(1, 3)))
  )

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=2)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16) #,
  # key=list(
  #   corner=c(0.05, 0.05),
  #   text=list(c("fixed effects only","fixed and random")),
  #   lines=list(lty=c(1, 3)))
  )

(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=2)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16) #,
  # key=list(
  #   corner=c(0.05, 0.05),
  #   text=list(c("fixed effects only","fixed and random")),
  #   lines=list(lty=c(1, 3)))
  )

@@

We proceed to fit several mixed-effects models to the data, using the `compareCoefs()` function from the \pkg{car} package [@FoxWeisberg:2019] to display the fixed-effect estimates for these models. We discuss each of these models below.
```{r fitted-mixed-models}
mod.0 <- lmer(y ~ 1 + (1 | group), Data)
mod.1 <- lmer(y ~ x + (1 | group), Data)
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
library("car", quietly=TRUE)
compareCoefs(mod.0, mod.1, mod.2, mod.3)
```

The initial mixed-effects model, \code{mod.0}, is a simple random-intercepts model. We obtain predictions from this model for the fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students). Predictions from \code{mod.0} for the first 10 students are shown in Figure \ref{fig:plot1} (b). The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean. Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.

Our next model, \code{mod.1}, includes a fixed intercept and the fixed effect of \code{x}, along with a random intercept. Predictions from this model appear in Figure \ref{fig:plot1} (c). The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}. We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod.1} to \code{mod.0}, but that cluster-based cross-validation will prefer \code{mod.0} to \code{mod.1}.

Our third model, \code{mod.2}, includes the "contextual effect" of $x$---that is, the cluster mean \code{xm}---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept. This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).

@@

Finally, \code{mod.3} directly estimates the model used to generate the data. As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}. The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar to those from \code{mod.2}.

@@

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")
```
A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors. There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

### 2024-01-15 scratch/notes_john/JSS-article/JSS-article-reduced-2.tex
compress JSS paper section on contrived data

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 801552
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.32751
bias-adjusted cross-validation criterion = 19.30935
95% CI for bias-adjusted CV criterion = (15.83674, 22.78196)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\hypertarget{example-cv-of-clusters-versus-individuals-using-contrived-data}{%
\subsection{Example: CV of clusters versus individuals using contrived
data}\label{example-cv-of-clusters-versus-individuals-using-contrived-data}}

@@

Figure \ref{fig:plot1} (a) shows a scatterplot the data for a
representative group of 10 (without loss of generality, the first 10) of
the 100 students, displaying the 95\% concentration ellipse for each
cluster:

@@

\caption[(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data]{(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data}\label{fig:plot1}
\end{figure}
\end{CodeChunk}

@@

The initial mixed-effects model that we fit to the data is a simple
random-intercepts model:

\begin{CodeChunk}
\begin{CodeInput}
R> # random intercept only:
R> mod.0 <- lmer(y ~ 1 + (1 | group), Data)
R> summary(mod.0)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: y ~ 1 + (1 | group)
   Data: Data

REML criterion at convergence: 2103.1

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.03514 -0.72645 -0.01169  0.78477  2.04377

Random effects:
 Groups   Name        Variance Std.Dev.
 group    (Intercept) 2.900    1.703   
 Residual             2.712    1.647   
Number of obs: 500, groups:  group, 100

Fixed effects:
            Estimate Std. Error t value
(Intercept)  10.0018     0.1855   53.91
\end{CodeOutput}
\end{CodeChunk}

We will shortly consider three other, more complex, mixed models.

We proceed to obtain predictions from the random-intercept model
(\code{mod.0}) for fixed effects alone, as would be used for
cross-validation based on clusters (i.e., students), and for fixed and
random effects---the BLUPs---as would be used for cross-validation based
on cases (i.e., occasions within students). Predictions for the
random-intercept model, \code{mod.0}, for the first 10 students are
shown in Figure \ref{fig:plot1} (b). The fixed-effect predictions for
the various individuals are identical---the estimated fixed-effects
intercept or estimated general mean of \(y\)---while the BLUPs are the
sums of the fixed-effects intercept and the random intercepts, and are
only slightly shrunken towards the general mean. Because in our
artificial data there is no population relationship between age and
skill, the fixed-effect-only predictions and the BLUPs are not very
different.

Our next model, \code{mod.1}, includes a fixed intercept and fixed
effect of \code{x} along with a random intercept:

\begin{CodeChunk}
\begin{CodeInput}
R> # effect of x and random intercept:
R> mod.1 <- lmer(y ~ x + (1 | group), Data)
R> summary(mod.1)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: y ~ x + (1 | group)
   Data: Data

REML criterion at convergence: 1564.5

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.90160 -0.63501  0.01879  0.55407  2.82932

Random effects:
 Groups   Name        Variance Std.Dev.
 group    (Intercept) 192.9406 13.8903 
 Residual               0.2569  0.5068 
Number of obs: 500, groups:  group, 100

Fixed effects:
             Estimate Std. Error t value
(Intercept) -33.91892    1.56446  -21.68
x             0.96529    0.01581   61.05

Correlation of Fixed Effects:
  (Intr)
x -0.460
\end{CodeOutput}
\end{CodeChunk}

Predictions from this model appear in Figure \ref{fig:plot1} (c). The
BLUPs fit the observed data very closely, but predictions based on the
fixed effects alone, with a common intercept and slope for all clusters,
are very poor---indeed, much worse than the fixed-effects-only
predictions based on the simpler random-intercept model, \code{mod.0}.
We therefore anticipate (and show later in this section) that case-based
cross-validation will prefer \code{mod1} to \code{mod0}, but that
cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.

Our third model, \code{mod.2}, includes the ``contextual effect'' of
\(x\)---that is, the cluster mean \code{xm}---along with \(x\) and the
intercept in the fixed-effect part of the model, and a random intercept:

\begin{CodeChunk}
\begin{CodeInput}
R> # effect of x, contextual (student) mean of x, and random intercept:
R> mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
R>         # equivalent to y ~ I(x - xm) + xm + (1 | group)
R> summary(mod.2)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: y ~ x + xm + (1 | group)
   Data: Data

REML criterion at convergence: 1169.2

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.98466 -0.63750  0.00191  0.55682  2.73246

Random effects:
 Groups   Name        Variance Std.Dev.
 group    (Intercept) 3.3986   1.8435  
 Residual             0.2552   0.5052  
Number of obs: 500, groups:  group, 100

Fixed effects:
            Estimate Std. Error t value
(Intercept)  9.47866    0.61705   15.36
x            0.99147    0.01597   62.07
xm          -0.97998    0.02055  -47.68

Correlation of Fixed Effects:
   (Intr) x     
x   0.000       
xm -0.600 -0.777
\end{CodeOutput}
\end{CodeChunk}

This model is equivalent to fitting
\code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that
generated the data once the coefficient of the contextual predictor
\code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).

@@

Finally, \code{mod.3} directly estimates the model used to generate the
data. As mentioned, it is a constrained version of \code{mod.2}, with
the coefficient of \code{xm} set to 0, and with \code{x} expressed as a
deviation from the cluster mean \code{xm}:

\begin{CodeChunk}
\begin{CodeInput}
R> # model generating the data (where Bb = 0)
R> mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
R> summary(mod.3)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: y ~ I(x - xm) + (1 | group)
   Data: Data

REML criterion at convergence: 1163.2

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.97703 -0.63204  0.00627  0.56032  2.72489

Random effects:
 Groups   Name        Variance Std.Dev.
 group    (Intercept) 3.3913   1.8415  
 Residual             0.2552   0.5052  
Number of obs: 500, groups:  group, 100

Fixed effects:
            Estimate Std. Error t value
(Intercept) 10.00176    0.18553   53.91
I(x - xm)    0.99147    0.01597   62.07

Correlation of Fixed Effects:
          (Intr)
I(x - xm) 0.000 
\end{CodeOutput}
\end{CodeChunk}

The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e),
are therefore similar to those from \code{mod.2}.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> Auto$cylinders <- factor(Auto$cylinders,
+                          labels=c("3.4", "3.4", "5.6", "5.6", "8"))
R> Auto$year <- as.factor(Auto$year)
R> Auto$origin <- factor(Auto$origin,
+                       labels=c("America", "Europe", "Japan"))
R> rownames(Auto) <- make.names(Auto$name, unique=TRUE)
R> Auto$name <- NULL
R> 
R> library("car", quietly=TRUE)
R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
+                   smooth=list(spread=FALSE), data=Auto, pch=".")
\end{CodeInput}
\begin{figure}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat){
  if (!all(y %in% c(0, 1))) stop("response values not all 0 or 1")
  if (any(yhat < 0) || any(yhat > 1)) stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x123a02cc0>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

>>>>

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 610656
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.39106
bias-adjusted cross-validation criterion = 19.36966
95% CI for bias-adjusted CV criterion = (15.86196, 22.87735)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\hypertarget{example-contrasting-cluster-based-and-case-based-cv}{%
\subsection{Example: Contrasting cluster-based and case-based
CV}\label{example-contrasting-cluster-based-and-case-based-cv}}

@@

Figure \ref{fig:plot1} (a) shows a scatterplot the data for a
representative group of 10 (without loss of generality, the first 10) of
the 100 students, displaying the 95\% concentration ellipse for each
cluster.

@@

\caption[(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data]{(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data}\label{fig:plot1}
\end{figure}
\end{CodeChunk}

@@

We proceed to fit several mixed-effects models to the data, using the
\texttt{compareCoefs()} function from the \pkg{car} package
\citep{FoxWeisberg:2019} to display the fixed-effect estimates for these
models. We discuss each of these models below.

\begin{CodeChunk}
\begin{CodeInput}
R> mod.0 <- lmer(y ~ 1 + (1 | group), Data)
R> mod.1 <- lmer(y ~ x + (1 | group), Data)
R> mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
R> mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
R> library("car", quietly=TRUE)
R> compareCoefs(mod.0, mod.1, mod.2, mod.3)
\end{CodeInput}
\begin{CodeOutput}
Calls:
1: lmer(formula = y ~ 1 + (1 | group), data = Data)
2: lmer(formula = y ~ x + (1 | group), data = Data)
3: lmer(formula = y ~ x + xm + (1 | group), data = Data)
4: lmer(formula = y ~ I(x - xm) + (1 | group), data = Data)

            Model 1 Model 2 Model 3 Model 4
(Intercept)  10.002 -33.919   9.479  10.002
SE            0.186   1.564   0.617   0.186

x                    0.9653  0.9915        
SE                   0.0158  0.0160

xm                          -0.9800        
SE                           0.0206

I(x - xm)                             0.992
SE                                    0.016

\end{CodeOutput}
\end{CodeChunk}

The initial mixed-effects model, \code{mod.0}, is a simple
random-intercepts model. We obtain predictions from this model for the
fixed effects alone, as would be used for cross-validation based on
clusters (i.e., students), and for fixed and random effects---the
BLUPs---as would be used for cross-validation based on cases (i.e.,
occasions within students). Predictions from \code{mod.0} for the first
10 students are shown in Figure \ref{fig:plot1} (b). The fixed-effect
predictions for the various individuals are identical---the estimated
fixed-effects intercept or estimated general mean of \(y\)---while the
BLUPs are the sums of the fixed-effects intercept and the random
intercepts, and are only slightly shrunken towards the general mean.
Because in our artificial data there is no population relationship
between age and skill, the fixed-effect-only predictions and the BLUPs
are not very different.

Our next model, \code{mod.1}, includes a fixed intercept and the fixed
effect of \code{x}, along with a random intercept. Predictions from this
model appear in Figure \ref{fig:plot1} (c). The BLUPs fit the observed
data very closely, but predictions based on the fixed effects alone,
with a common intercept and slope for all clusters, are very
poor---indeed, much worse than the fixed-effects-only predictions based
on the simpler random-intercept model, \code{mod.0}. We therefore
anticipate (and show later in this section) that case-based
cross-validation will prefer \code{mod.1} to \code{mod.0}, but that
cluster-based cross-validation will prefer \code{mod.0} to \code{mod.1}.

Our third model, \code{mod.2}, includes the ``contextual effect'' of
\(x\)---that is, the cluster mean \code{xm}---along with \(x\) and the
intercept in the fixed-effect part of the model, and a random intercept.
This model is equivalent to fitting
\code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that
generated the data once the coefficient of the contextual predictor
\code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).

@@

Finally, \code{mod.3} directly estimates the model used to generate the
data. As mentioned, it is a constrained version of \code{mod.2}, with
the coefficient of \code{xm} set to 0, and with \code{x} expressed as a
deviation from the cluster mean \code{xm}. The predictions from
\code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar
to those from \code{mod.2}.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> Auto$cylinders <- factor(Auto$cylinders,
+                          labels=c("3.4", "3.4", "5.6", "5.6", "8"))
R> Auto$year <- as.factor(Auto$year)
R> Auto$origin <- factor(Auto$origin,
+                       labels=c("America", "Europe", "Japan"))
R> rownames(Auto) <- make.names(Auto$name, unique=TRUE)
R> Auto$name <- NULL
R> 
R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
+                   smooth=list(spread=FALSE), data=Auto, pch=".")
\end{CodeInput}
\begin{figure}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat){
  if (!all(y %in% c(0, 1))) stop("response values not all 0 or 1")
  if (any(yhat < 0) || any(yhat > 1)) stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x139564668>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}
